 

### 2021-01-15 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves that will potentially reduce [forest] biomass acculumation and disrupt photosynthesis and metabolic processes across plant species [O'Sullivan;REF ;@perkinsIncreasingFrequencyIntensity2012].
These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].

@@

The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
Increasing temperature has profound impacts on canopy cover, such that temperate and tropical forest canopies are already or will soon be functioning beyond their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Additionally, increased temperature is driving more frequent droughts, causing hydraulic limitation in forests [ref]. Water limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to increased sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020].Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration, thereby buffering understory microclimates [REF].

@@

While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in overstory and understory is still unclear. Thus, it is becoming more important than ever to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy. This will provide a pathway for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.

@@

Leaf traits are shaped by light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, hea, C uptake and metabolism by leaves across forest vertical strata [Rosendaal et al. 2006]. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [Keenan and Niinemets et al.2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013]. As a result, degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also influences leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al. 2016].

@@

Leaf traits (functional and physiological) co-vary with their environment to maintain optimum temperature for metabolic processes under various environmental circumstances [@michaletzEnergeticCarbonEconomic2016].
Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy compared to shorter understory trees that are shaded by overstory. As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and greater lobation than shade leaves
[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].

@@

Leaf temperature is generally associated to increase with leaf area ($LA$) [ Wright et al. 2019]. However, thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn traspirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in deterimining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension is generally greater in shaded understory leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration break the entirity of leaf dimension by enhancing a smaller characteristic dimension, all facilitate in dissapating heat more efficiently. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al. 2006].

- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).

@@

**Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. 3).
Leaf traits are regulated to varying biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015] 
$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].

@@

In tropical and temperate forests, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. **2?**).**
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999).Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015]. 
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

@@

*Stomatal conductance*
Photosynthesis can be largely constrained by stomatal conductance. 
-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al. 2019)

@@

**Leaf thermal tolerance (Tcrit/ T50)**
- define.
Typical values ~45-50$^\circ$C, varying somewhat 
across latitude/climate (O’sullivan et al. 2017),
with elevation [@feeley_thermal_2020;  Slot unpublished data],
with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
- Elsa: also see: 
- Perez, T.M. and Feeley, K.J., 2020. Photosynthetic heat tolerances and extreme leaf temperatures. Functional Ecology, 34(11), pp.2236-2245.
- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.

@@

>>>>

Globally, temperatures have risen # $^\circ$C since YEAR and are expected to rise an additional #-#$^\circ$C by YEAR (IPCC REF). 
There is also increasing severity and frequency of heat waves that will potentially reduce [forest] biomass acculumation, and disrupt photosynthesis and metabolic processes across plant species [Sullivan et al. 2020, @perkinsIncreasingFrequencyIntensity2012].
These changes are expected to have profound effects on forest biodiversity and ecosystem function (REFS), and the resulting feedbacks to the climate system (e.g., decreased C storage, altered albedo and hydrology) will in turn have a critical influence on the future of trajectory of climate change [@bonan_ecological_2016, Bonan 2008].

@@

The capacity of buffering largely depends on canopy cover and water availability [@davis_microclimatic_2019] each of which is subjected to change through climate-driven disturbances such as drought, deforestation, fire and related disruptions [@senfCanopyMortalityHas2018a].
Increasing global temperatures can increase canopy leaf temperatures sometimes above ambient temperature when transpiration is low, leading to temperate and tropical forest canopies functioning beyond or close to their optimal photosynthetic threshold [@mau_temperate_2018; @huangAirTemperatureOptima2019]. Conversly, available water facilitates leaf transpiration through open stomata, however increased relative humdity in tropical regions indirectly caused by increased temperature, reduces transpiration leading to high leaf temperatures in the canopy [Tibbitts 1979; Perez and Feeley 2018].

Frequent droughts related hydraulic limitation in forests alters species thermoregulation [@sastry_leaf_2018], and also decreases tree canopy circumference [@aussenacInteractionsForestStands2000a] due to sensitivity of overstory to variability in water availability and precipitation [@rollinsonClimateSensitivityUnderstory 2020]. Larger trees suffer hydraulic stress during drought [resulting in heat stressed canopy-leaves], which may contribute to increasing larger tree mortality in forests around the world should frequencies of drought continue to increase [@mcdowellMechanismsPlantSurvival2008; @allenUnderestimationGlobalVulnerability2015, Brienen et al. 2015; @bennettLargerTreesSuffer2015; @stovallTreeHeightExplains2019a]. Increases in loss of canopy cover [@senfCanopyMortalityHas2018a; @senfMappingForestDisturbance2020] are also associated with reduced canopy structural complexity, altering microclimates with local heating effects and surface energy balance components [@zellweger_forest_2020; @starkReframingTropicalSavannization2020]. However, in mesic forest conditions, availability of soil water enables canopies with suitable foliar traits to remain cool through continuous evapotranspiration even during increasing air temperature, thereby buffering understory microclimates [REF].

@@

While there is evidence for temperature driven shifts in forest dynamics, leaf-level metabolic and performance changes influenced by biophysical conditions and warming in the overstory and understory is still unclear. Thus, it is becoming important to understand foliage thermal sensitivity in forests along the vertical canopy profile, from the understory to the top of the canopy. This will provide a pathway for how rising temperature affects forest ecology and biodiversity, energy balance, ecosystem function, and biosphere-atmosphere interactions with additional feedbacks to forest structure and microclimates.

@@

Leaf traits co-vary with their environment [@michaletzEnergeticCarbonEconomic2016] and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light, optimizing water uptake, heat, C uptake and metabolism across forest vertical strata [Rosendaal et al. 2006]. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabilize in woody and herbaceous species [Keenan and Niinemets et al.2016; @niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves, such as phytochromes, mediate leaf-level responses to PAR, and the ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013]. As a result, degrees of high irradiance give rise to variation in sun-leaf traits, while degrees of low irradiance-- characterized by shade-- result in variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also influences leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al. 2016].

@@

Taller trees in dense forests tend to be exposed to greater irradiance but maintain shaded leaves in canopy interior and lower canopy compared to shorter understory trees that are shaded by overstory. As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and greater lobation than shade leaves
[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].

@@

Leaf temperature is generally associated to increase with leaf area ($LA$) [ Wright et al. 2019]. Thermal imagery shows that though leaf area can be a predictor for leaf temperatures ranges [$Trange$] across the surface of the leaf, effective leaf width is a better predictor of leaf temperature changes and leaf thermal time constant [@leighInfluenceLeafSize2017]. With increase in efffective leaf width, boundary layer thickness and resistance increases along with stomtal resistance [@leighInfluenceLeafSize2017; Wright et al. 2019; @fauset_differences_2018].This lowers evaporation from stomatal pores, and in turn traspirational cooling in large leaves [@fauset_differences_2018]. Because the rate of heat convection per unit area between leaf and air is lower for large leaves and greater for small leaves, small leaves are able to dissapate heat more efficiently than large leaves [@leighInfluenceLeafSize2017]. Using leaf energy balance to compare leaf temperatures of two species $Miconia$ and $Alchornea$, Fauset et al. 2016 found leaf width to be a dominant factor in deterimining changes in leaf temperature, which was evidently greater for large width leaves of $Miconia$ compared to $Alchornea$. Leaf width and characteristic dimension is generally greater in shaded leaves along the vertical gradient than exposed upper canopy [ref]. Additionally, leaf characteristics such as lobes, leaf elongation and serration break the entirity of leaf dimension by enhancing a smaller characteristic dimension, all facilitate in dissapating heat more efficiently. Leaf lobes are deeper in exposed canopy than in shaded understory, as well as leaf elongation and serration are more pronounced in sun exposed canopies [Vogel 1970; Roth-Nebelsick 2001;Schuepp 1993; Sack et al. 2006].However, Perez and Feeley et al. 2020 showed only a slight negetive effect on leaf temperature with leaf width.

@@

**Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. 3).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf, and thus leaf temperatures are decoupled from air temperature through adaptive physiological traits [@michaletzPlantThermoregulationEnergetics2015; Perez et al. 2020] 
$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability increases [@michaletzPlantThermoregulationEnergetics2015].

@@

In a tropical forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. **2?**).**
[Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999). 
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

@@

*Stomatal conductance*
Photosynthesis can be largely constrained by stomatal conductance. 
-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al. 2019)
- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).

@@

**Leaf thermal tolerance (Tcrit/ T50)**
- define.
Typical values ~45-50$^\circ$C, varying somewhat 
across latitude/climate (O’sullivan et al. 2017),
with elevation [@feeley_thermal_2020;  Slot unpublished data],
with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
Contrary to O'Sullivan et al. 2017's findings, Perez and Feeley et al. 2020 found that across inter-species sun leaves, higher photosynthetic heat tolerance in leaves lead to hotter leaf temperatures and narrower thermal safety margins and thus may be at a greater risk of thermal damage.  
[- Elsa: also see: 
- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.]

Supporting the significant negetive relationship between $T50$ and thermal safety margins, photosynthesis [PS2] temperature regulation has been associated leaf thermal tolerance in order to maximize net carbon assimilation. However, randomly collected within-canopy leaves of individual mature plants in Florida showed that Tmax is not directly coordinated with high PS2 heat tolerance and does not necessarily increase leaf thermal tolerance [shows no direct correlation]; nor increases carbon assimilation at higher temperature. Rather, high PS2 heat tolerance reduces optimum temperature for C assimilation [$Topt$] because high PS2 heat tolerance requires a greater cost of PS2 metabolic activity maintainance thus not contributing to increased C assimilation. Species exhibiting low carbon assimilation rate tended to have greater stress tolerance. In addition, species with low stomtal conductance exhibit greater PS2 heat tolerance because earlier closure of stomata ceases C assimaltion before thermal limits are reached [Perez et al. 2020]

@@

increased irradiance = increased VPD = increased air temp = midday stomatal closure =  decreased latent heat loss = increase leaf temperature = decreased C assimilation = greater PS2 heat tolerance = optimum rate of photosynthesis decreases = exhibiting lower t opt (optimum temp for c assimilation)

incr.humidity = decreased VPD = dec. latent heat loss = limited c assimilation [Perez et al. 2020]
